---
title: "Getting Data for Segregation Measures from the Census API"
author: "Angel Mendiola Ross"
date: "5/8/2020"
output: html_document
---

```{r setup}
knitr::opts_knit$set(root.dir = '/Users/angelmr/ps239t-final-project/')
```

## 1.0 Setup

```{r}

### Clear global environment
rm(list=ls())

# Set working directory
getwd()

library(pacman)
pacman::p_unload(all)

pacman::p_load(
  tidyverse, #dplyr, readr, etc.
  plyr, #for editing factor levels
  magrittr, #%<>% operator
  readxl, #for excel workbooks
  tidyr, #to transform from wide to long
  ggplot2, #for plotting
  tidycensus #census api
  #purrr #for map function
)

#census api key
census_api_key("1ce78664b294208f88f53c49746f378a51893b62")

```

## 1.1. Preparing Census API  call

The purpose of this script is to call the data I need to create segregation measures, one of the main explanatory variables for my analysis. I am creating indices of dissimilarity between homeowners and renters as well as Black-white and Latino-white dissimilarity indices. 

Following Trounstine (2018), in the future, I would also like to compute entropy scores and construct the Theil's H index.

To start, I need to get the data necessary at the census tract level. I tried to run this analysis at the block group level because some of the cities in my sample are small and census tracts range from 2,500 to 8,000 people. As a result, some of the cities are comprised of only one tract. The block group (which ranges from 600 to 3,000 people) would be an interesting alterantive, but tidycensus is not working with block groups after 2009. So I am sticking with the tract level at this time.

```{r}

# the years I want data for
years <- lst(2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018)

# the main variables I need (household-level data for tenure and people-level data for race/ethnicity)
tenure_tract <- c(
  total_hhs = "B25003_001",
  owner_hhs = "B25003_002",
  renter_hhs = "B25003_003",
  total_pop = "B03002_001",
  nhw = "B03002_003",
  nhb = "B03002_004",
  nha = "B03002_006",
  his = "B03002_012")

```

## 1.2 Calling the tract-level data

``` {r}

# this function generates a dataframe with the variables of interest by tract from 2009-2018
tenure_by_tract <- map_dfr(
  years,
  ~ get_acs(
    geography = "tract",
    variables = tenure_tract,
    state = "CA",
    year = .x,
    survey = "acs5"
  ),
  .id = "year"
)

# deleting moe from acs dataframe because I do not need it for the analysis I am conducting
tenure_by_tract %<>%
  select(-c(moe))

tenure_by_tract %>% nrow() #rows in dataset=636496
tenure_by_tract %>% distinct() %>% nrow() #unique rows in dataset=636496

#using spread to transform the estiamte rows into columns
tenure_by_tract <- tenure_by_tract %>%
  spread(variable, estimate)

```

## 1.2 Now calling the place-level data

I will need to join the tract-level data with the place-level data to be able to construct the dissimilarity indices by place.

``` {r}

# calling tenure data at the place-level
tenure_by_place <- map_dfr(
  years,
  ~ get_acs(
    geography = "place",
    variables = tenure_tract,
    state = "CA",
    year = .x,
    survey = "acs5"
  ),
  .id = "year"
)

# deleting moe from acs dataframe
tenure_by_place %<>%
  select(-c(moe))

tenure_by_place %>% nrow() #rows in dataset=117992
tenure_by_place %>% distinct() %>% nrow() #unique rows in dataset=117992

#using spread to transform variable rows to columns
tenure_by_place <- tenure_by_place %>%
  spread(variable, estimate)

```

## 1.3 Now calling the metropolitan area data

I need tract-level and place-level data for the dissimilarity indices, but I also need the metro area data to compute the entropy and Theil's H index (Trounstine 2018). So I'm getting that data for the metro area as well.

```{r}

# For some reason, when I include the state argument, I get the following API call error: unknown/unsupported geography hierarchy

# But if I run it for the whole country, it works. So I will have to run it for the whole country and subset after
tenure_by_metro <- map_dfr(
  years,
  ~ get_acs(
    geography = "metropolitan statistical area/micropolitan statistical area",
    variables = tenure_tract,
    year = .x,
    survey = "acs5"
  ),
  .id = "year"
)

# subsetting to CA only
tenure_by_metro %<>%
  subset(grepl(", CA", NAME))

# checking MSAs
unique(tenure_by_metro$NAME)
# looks good

tenure_by_metro %>% nrow() #rows in dataset=2752
tenure_by_metro %>% distinct() %>% nrow() #unique rows in dataset=2752

# adding a bool for whether the MSA is a metro area (true) or micro area (false)
tenure_by_metro %<>%
  mutate(metro = ifelse(grepl("Metro", NAME), TRUE, FALSE))

# checking random sample
tenure_by_metro %>%
  sample_n(size = 10)

# deleting moe from acs dataframe
tenure_by_metro %<>%
  select(-c(moe))

#using spread to transform variable rows to columns
tenure_by_metro <- tenure_by_metro %>%
  spread(variable, estimate)

```

## 1.4 Rearranging columns and saving out data files

``` {r}
# checking columns
names(tenure_by_metro)
names(tenure_by_place)
names(tenure_by_tract)

# saving in correct order
tenure_by_metro <- tenure_by_metro[c(1:4,11,9,10,12,8,7,6,5)]
tenure_by_place <- tenure_by_place[c(1:3,10,8,9,11,7,6,5,4)]
tenure_by_tract <- tenure_by_tract[c(1:3,10,8,9,11,7,6,5,4)]

#saving raw tenure/race data for tract, place, and metro
save(tenure_by_tract,
     file="data/raw/tenure_by_tract_raw.RData")

save(tenure_by_place,
     file="data/raw/tenure_by_place_raw.RData")

save(tenure_by_metro,
     file="data/raw/tenure_by_metro_raw.RData")
```
