---
title: "Computing Segregation Measures"
author: "Angel Mendiola Ross"
date: "5/8/2020"
output: html_document
---

```{r setup}
knitr::opts_knit$set(root.dir = '/Users/angelmr/ps239t-final-project/')
```

## 1.0 Setup

```{r}

### Clear global environment
rm(list=ls())

# Set working directory
getwd()

library(pacman)
pacman::p_unload(all)

pacman::p_load(
  tidyverse, #dplyr, readr, etc.
  magrittr, #%<>% operator
  tidyr, #to transform from wide to long
  ggthemes # for pretty charts
)

```

The goal of this script is to take the data generated from the Census API and compute a series of segregation measures, beginning with the index of dissimilarity for renters.

## 1.1 Load files

The first step is to load the files I generated from the Census API as well as the crosswalk from the LODES dataset.

I downloaded the Census crosswalk from the following directory: <https://lehd.ces.census.gov/data/lodes/LODES7/ca/>. It includes information down to the block level. Because it is a pretty large file, I subsetted it down to California and the geographies needed. That file loads as xwalk_small.


```{r}

# Loading the necessary files 
load("data/raw/tenure_by_tract_raw.RData")
load("data/raw/tenure_by_place_raw.RData")
load("data/raw/tenure_by_metro_raw.RData")
load("xwalks/census_xwalk_small.RData")

# the crosswalk still requires some wrangling
names(xwalk_small)

# dropping the block-level data
xwalk_small <- xwalk_small[c(5,6,11,12,3,4,9,10)]

xwalk_small %>% nrow() #rows in dataset=710145
xwalk_small %>% distinct() %>% nrow() #unique rows in dataset=12500

# deleting duplicate rows based on trct (which is the census tract)
xwalk_small %<>% distinct(trct, .keep_all = TRUE)

xwalk_small %>% distinct() %>% nrow() #unique rows in dataset=8059 and there are 8,057 tracts in CA so looks good

```

## 1.2 Joining Datasets

``` {r}

# prepping tract-level census data for join on trct in crosswalk
tenure_by_tract %<>% dplyr::rename(trct = GEOID)

# Starting tracts as the base dataset, but need to subset for ACS years 2010-2018 because the 2009 tract data is in the 2000 boundaries
renter_seg <- tenure_by_tract %>% subset(year>2009)

# Add columns with data from the tract-place crosswalk, matching on 'GEOID' variable 
renter_seg <- left_join(x = renter_seg,
                          y = xwalk_small,
                          by = "trct")

renter_seg %>% nrow() #rows in dataset=72513
renter_seg %>% distinct() %>% nrow() #unique rows in dataset=72513

# renaming geoid for place df "stplc" which is how it is in the crosswalk
tenure_by_place %<>% dplyr::rename(stplc = GEOID)

# Add columns with place-level variables to renter_seg, matching on 'year' & 'PLACE_GEOID' variable 
renter_seg <- left_join(x = renter_seg,
                          y = tenure_by_place,
                          by = c('year', 'stplc'))


renter_seg %>% nrow() #rows in dataset=72513
renter_seg %>% distinct() %>% nrow() #unique rows in dataset=72513

# renaming geoid in the metro data to "cbsa"
tenure_by_metro %<>% dplyr::rename(cbsa = GEOID)

# chaning cbsa to a character variable in the renter_seg dataframe
renter_seg$cbsa <- as.character(renter_seg$cbsa)

# Add columns with metro-level variables to renter_seg, matching on 'year' & 'PLACE_GEOID' variable 
renter_seg <- left_join(x = renter_seg,
                          y = tenure_by_metro,
                          by = c('year', 'cbsa'))

```


## 1.3 Cleaning the Merged Renter Segregation Dataframe

``` {r}

# first step is viewing names and assigning variable names that make sense
names(renter_seg)

# starting with renaming renter variables
renter_seg %<>% dplyr::rename(renters_tract = renter_hhs.x)
renter_seg %<>% dplyr::rename(renters_place = renter_hhs.y)
renter_seg %<>% dplyr::rename(renters_metro = renter_hhs)

# now renaming homeowner variables
renter_seg %<>% dplyr::rename(owners_tract = owner_hhs.x)
renter_seg %<>% dplyr::rename(owners_place = owner_hhs.y)
renter_seg %<>% dplyr::rename(owners_metro = owner_hhs)

# renaming total households variables
renter_seg %<>% dplyr::rename(hhs_tract = total_hhs.x)
renter_seg %<>% dplyr::rename(hhs_place = total_hhs.y)
renter_seg %<>% dplyr::rename(hhs_metro = total_hhs)

# moving on to demographic variables
renter_seg %<>% dplyr::rename(total_pop_tract = total_pop.x)
renter_seg %<>% dplyr::rename(total_pop_place = total_pop.y)
renter_seg %<>% dplyr::rename(total_pop_metro = total_pop)

# non-hispanic white variable
renter_seg %<>% dplyr::rename(nhw_tract = nhw.x)
renter_seg %<>% dplyr::rename(nhw_place = nhw.y)
renter_seg %<>% dplyr::rename(nhw_metro = nhw)

# non-hispanic black variable
renter_seg %<>% dplyr::rename(nhb_tract = nhb.x)
renter_seg %<>% dplyr::rename(nhb_place = nhb.y)
renter_seg %<>% dplyr::rename(nhb_metro = nhb)

# non-hispanic asian variable
renter_seg %<>% dplyr::rename(nha_tract = nha.x)
renter_seg %<>% dplyr::rename(nha_place = nha.y)
renter_seg %<>% dplyr::rename(nha_metro = nha)

# and hispanic/latino variable
renter_seg %<>% dplyr::rename(his_tract = his.x)
renter_seg %<>% dplyr::rename(his_place = his.y)
renter_seg %<>% dplyr::rename(his_metro = his)

# now changing the "NAME" variables
renter_seg %<>% dplyr::rename(tract_name = NAME.x)
renter_seg %<>% dplyr::rename(place_name = NAME.y)
renter_seg %<>% dplyr::rename(cbsa_name = NAME)

# rearranging the columns so that the geo columns correspond to the data better
renter_seg <- renter_seg[c(1:14,19:27,15:18,28:37)]

# saving combined renter_seg data
save(renter_seg,
     file="data/renter_seg.RData")

```


## 1.4 Computing Renter Segregation (index of dissimilarity)

Now that the dataset is cleanly labeled and ordered, I will begin computing the variables necessary for the index of dissimilarity for renters-homeowners.

The formula for the index of dissimilarity that I use is here is based on Massey and Denton's (1988) "The Dimensions of Residential Segregation." It is also described in the Census Bureau's Measures of Segregation appendix: <https://www.census.gov/hhes/www/housing/resseg/pdf/app_b.pdf>. (see top right of page 122)

``` {r}

## NOTE: I use the dplyr:: prefix in the following chunk because group_by does not work without it due to plyr

# computing the necessary renter proportions for tracts and for cities
renter_seg %<>%
  dplyr::mutate(prenter_tract = renters_tract/hhs_tract)

renter_seg %<>%
  dplyr::mutate(prenter_city = renters_place/hhs_place)

# computing absolute difference of the two proportions
renter_seg %<>%
  dplyr::mutate(prenter_abs_diff = abs(prenter_tract - prenter_city))
                
# computing the numerator of the index of dissimilarity (based on city-level data)
renter_seg %<>%
  dplyr::group_by(place_name, year) %>%
  dplyr::mutate(diss_num_city = sum(hhs_tract*prenter_abs_diff, na.rm=TRUE)) %>% 
  dplyr::ungroup()

# assign NAs if there is no city-level data
renter_seg %<>%
  dplyr::mutate(diss_num_city = ifelse(is.na(prenter_city), NA, diss_num_city))

# computing T
renter_seg %<>%
  dplyr::group_by(place_name, year) %>%
  dplyr::mutate(T_diss = sum(hhs_tract, na.rm=TRUE)) %>% 
  dplyr::ungroup()

# computing X
renter_seg %<>%
  dplyr::group_by(place_name, year) %>%
  dplyr::mutate(X_diss = sum(renters_tract, na.rm=TRUE)) %>% 
  dplyr::ungroup()

# computing P (ratio of X to T)
renter_seg %<>%
  dplyr::mutate(P_diss = X_diss / T_diss)

# adding a variable for the number of tracts in each city
renter_seg %<>%
  dplyr::group_by(place_name, year) %>%
  dplyr::mutate(n_tracts = n())

# checking to see if the large differences between the cities and tracts are due to cities with only a small number of tracts within in
renter_seg %>%
  dplyr::mutate(diff_P_city = P_diss - prenter_city) %>%
  subset(diff_P_city>.30) %>%
  dplyr::select("year","place_name","prenter_city","P_diss","diff_P_city","n_tracts")

renter_seg %>%
  dplyr::mutate(diff_P_city = P_diss - prenter_city) %>%
  subset(diff_P_city<(-.30)) %>%
  dplyr::select("year","place_name","prenter_city","P_diss","diff_P_city","n_tracts")
# The only big issues here are the city of Industry and Rancho Santa Margarita

# computing the dissimilarity denominator (based on aggregated tract-level data): (2PT)(1-P) 
renter_seg %<>%
  dplyr::group_by(place_name, year) %>%
  dplyr::mutate(diss_denom = 2*(P_diss*T_diss)*(1-P_diss)) %>% 
  dplyr::ungroup()

# computing the numerator of the index of dissimilarity (based on aggregated tract-level data): sum(hhs_tract)*abs(prenter_tract-P_diss)

# first computing pi_abs_diff
renter_seg %<>%
  dplyr::mutate(pi_abs_diff = abs(prenter_tract - P_diss))

# now the numerator
renter_seg %<>%
  dplyr::group_by(place_name, year) %>%
  dplyr::mutate(diss_num = sum(hhs_tract*pi_abs_diff, na.rm=TRUE)) %>% 
  dplyr::ungroup()

# computing the dissimilarity denominator (based on city-level data): (2PT)(1-P) 
renter_seg %<>%
  dplyr::group_by(place_name, year) %>%
  dplyr::mutate(diss_denom_city = 2*(hhs_place*prenter_city)*(1-prenter_city)) %>% 
  dplyr::ungroup()

# now finally computing the renter-owner index of dissimilarity

# based on city-level data
renter_seg %<>%
  dplyr::mutate(diss_index_city = diss_num_city / diss_denom_city)

# based on aggregated tract-level data
renter_seg %<>%
  dplyr::mutate(diss_index = diss_num / diss_denom)

```

## 1.5 Transforming Data Frame into City-level Data

Even though this is only for the 2010 to 2018 data and we would not expect much variation in such a short time frame for something like housing, I want to see if there is any variation.

I have my data right now organized by census tract. But now that I computed the dissimilarity index, I want it organized by place (or city). So, first I will create a new dataframe with cities as the main unit of analysis and compute some summary statistics.

``` {r}
# stripping data frame down to just cities
renter_seg_city <- distinct(renter_seg, year, stplc, .keep_all = TRUE)

# looking at summary stats
renter_seg_city %>%
  subset(!is.na(cbsa) & !is.na(place_name)) %>%
  select(n_tracts) %>%
  summary(n_tracts)

# assigning as NA the 27 values that are infinity
renter_seg_city %<>%
  dplyr::mutate(diss_index_city = ifelse(diss_index_city>10000, NA, diss_index_city))

# now I can see the true mean for the dissimilarity index based on city-level variables
renter_seg_city %>%
  subset(!is.na(cbsa) & !is.na(place_name)) %>%
  select(diss_index, diss_index_city, n_tracts) %>%
  summary(diss_index)
# there are some values that are very high on the city-level index that are skewing the mean

# if I subset to cities/places with diss_index_city is less than 1, the means and medians are much closer
renter_seg_city %>%
  subset(!is.na(cbsa) & !is.na(place_name) & diss_index_city<1) %>%
  select(diss_index, diss_index_city, n_tracts) %>%
  summary(diss_index)

```

## 1.6 Plotting the Dissimilarity Index Over Time

To get a feel for my data and prepare for plotting, I will transform the year variable (which was coerced into a character during an earlier join) into a numeric and compute yearly means.
```{r}
renter_seg_city$year <- as.numeric(as.character(renter_seg_city$year))

# creating summary dfs to plot
means <- renter_seg_city %>% 
  group_by(year) %>%
  summarise(diss_index = mean(diss_index, na.rm = T))

means_city <- renter_seg_city %>% 
  group_by(year) %>%
  summarise(diss_index_city = mean(diss_index_city, na.rm = T))

# checking yearly means
means
means_city

```


Now I can plot renter segregation based on both indices of dissimilairy. This one is using the aggregated tract-level measure (rather than actual city-level data):
``` {r, echo=F}
ggplot(renter_seg_city, aes(x=year, y=diss_index)) +
  geom_point() +
  geom_line(data = means, alpha = .8, size = 1, color = "orange") +
  ggtitle("Renter Segregation, California") +
  labs(y="Index of Dissimilarity (aggregated from tract-level measure)", x = "Year") +
  theme_economist() + 
  scale_colour_economist()

# Save plot 
ggsave(filename="plots/renter_segregation.pdf")
```

And this one is using the actually city-level data. Note that the index is only suppose to range from 0 to 1 but because this is based on the city-level data rather than tract aggregates, there are some cities where the index is greater than 1 because census tract boundaries are not cleanly divided by cities.
``` {r, echo=F}
ggplot(renter_seg_city, aes(x=year, y=diss_index_city)) +
  geom_point() +
  geom_line(data = means_city, alpha = .8, size = 1, color = "orange") +
  ylim(0,3) +
  ggtitle("Renter Segregation, California") +
  labs(y="Index of Dissimilarity (from city-level data)", x = "Year") +
  theme_economist() + 
  scale_colour_economist() 

# Save plot 
ggsave(filename="plots/renter_segregation_city.pdf")
```

## 1.7 Saving Renter Segregation Data

I don't need all variables in the dataframe for my main policing analysis, so here I am reducing the file down a bit and saving to an RData file.

```{r}

# checking names
names(renter_seg_city)

# selecting necessary variables
renter_seg_city <- renter_seg_city[c(1,13:37,45,51,50)]

# saving to data folder
save(renter_seg_city,
     file = "data/renter_seg_city.RData")

```
